---
title: "Getting started with ivctools: diagnosing single-instrument exogeneity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ivctools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

## 1. When is this package for you?

`ivctools` is for a specific but common situation in education and the social
sciences: you have a **pretest--posttest design**, you want to estimate the
causal effect of a treatment, and you are relying on a **single instrumental
variable** whose exogeneity you cannot fully defend.

You should consider the IVC diagnostic if you can identify four things in your
data:

- an **outcome** measured after treatment (the posttest, `Y`),
- a **pretest** measured before treatment (`P`),
- a **treatment** indicator or dose (`A`),
- a candidate **instrument** (`Z`) that you hope affects `Y` only through `A`.

Standard overidentification tests (Sargan, Hansen J) cannot help here: they
need *two or more* instruments. The IVC diagnostic instead uses the pretest as
a built-in benchmark, so it works with a single instrument.

```{r setup}
library(ivctools)
```

## 2. The idea in one paragraph

If the instrument `Z` is valid, two different estimates of the treatment effect
should agree: the conventional IV estimate (`tau_IV`) and a *compass-corrected*
estimate built from the pretest (`tau_comp`). Their difference,
`Delta = tau_IV - tau_comp`, has expectation zero under a valid instrument. A
confidence interval for `Delta` that **excludes zero** is evidence that `Z`
violates exogeneity. Crucially, `tau_comp` remains a valid estimate of the
effect even when `Z` is invalid, so the gap between the two is informative.

## 3. A first analysis

The package ships with a small simulated dataset, `ivc_example`, which has a
mild exogeneity violation built in.

```{r data}
data(ivc_example)
head(ivc_example)
```

Run the diagnostic with one call. Give it the data frame and the column names.

```{r first-fit}
fit <- compare_iv_cv(ivc_example,
                     outcome = "Y", pretest = "P",
                     treat = "A", instrument = "Z",
                     se = "mom")
fit
```

## 4. Reading the output

Each quantity now reports its own standard error:

- **`tau_IV`** -- the conventional IV estimate. Biased if `Z` is invalid.
- **`tau_comp`** -- the compass-corrected estimate. Your best point estimate of
  the effect whether or not `Z` is valid.
- **`delta_hat`** -- the loading ratio used in the correction.
- **`Delta`** and its CI -- the diagnostic itself. If the CI excludes 0, the
  conclusion line flags evidence against exogeneity.

You can pull any value out directly; the fitted object is a plain list.

```{r extract}
fit$tau_comp
fit$se_tau_comp
fit$Delta
fit$ci
fit$reject
```

A picture often communicates the diagnostic best. The plot is a forest plot:
each of `tau_IV`, `tau_comp`, and `Delta` gets its own point and confidence
interval on a common scale, so you can see at a glance whether the two
estimates' intervals overlap. The `Delta` row is shaded and colored red when
its interval excludes zero.

```{r plot}
plot(fit)
```

## 5. Individual estimates and their standard errors

If you only want one estimate with its standard error, use the single-estimate
functions with `se = TRUE`.

```{r individual}
estimate_iv(ivc_example, "Y", "P", "A", "Z", se = TRUE)
estimate_cv(ivc_example, "Y", "P", "A", "Z", return_delta = TRUE, se = TRUE)
```

## 6. Covariates and survey weights

Real data usually need controls and, for complex samples, weights. Supply them
by column name.

```{r covariates, eval = FALSE}
compare_iv_cv(my_data,
              outcome = "posttest", pretest = "pretest",
              treat = "catholic", instrument = "z",
              covariates = c("female", "ses", "region"),
              weights = "survey_weight",
              se = "mom")
```

Covariates are handled by Frisch--Waugh--Lovell partialling: the outcome,
pretest, treatment, and instrument are residualized on the covariates before
the estimators are applied. With neither `covariates` nor `weights`, the
results reduce **exactly** to the covariate-free forms.

## 7. Multilevel data: cluster-robust inference

Educational data are often nested: students within schools, employees within
firms. If the treatment (or an omitted factor correlated with it) operates at
the cluster level, observations within a cluster are not independent, and
standard errors computed under an independence assumption can be badly
understated.

Pass `cluster = ` (a column name or a vector of cluster IDs) to `estimate_cv()`
or `compare_iv_cv()` to get cluster-robust standard errors instead. The point
estimates never change -- only how the standard error is computed.

```{r cluster-sim}
set.seed(1)
n_school <- 40
m        <- 25
school   <- rep(seq_len(n_school), each = m)
n        <- length(school)

Zs <- rnorm(n_school)                            # school-level component of Z
Us <- rnorm(n_school)                            # school-level confound
Gs <- rbinom(n_school, 1, plogis(1.5 * Zs))      # Z predicts school-level adoption
A  <- Gs[school]
U  <- Us[school] + rnorm(n)                       # confound: school part + individual part
Z  <- Zs[school] + 0.5 * rnorm(n)
P  <- 1.2 * U + rnorm(n)
Y  <- 0.5 * A + 1.2 * U + Us[school] + rnorm(n)   # an extra school-level factor in Y
d_school <- data.frame(A, Z, P, Y, school)

# Ignoring the school structure understates the SE here, because both the
# treatment and part of the outcome operate at the school level.
estimate_cv(d_school, "Y", "P", "A", "Z", se = TRUE)
estimate_cv(d_school, "Y", "P", "A", "Z", se = TRUE, cluster = "school")
```

`compare_iv_cv()` accepts the same argument for both `se = "mom"` (a
cluster-robust sandwich) and `se = "bootstrap"` (whole clusters are
resampled together rather than individual rows):

```{r cluster-compare}
compare_iv_cv(d_school, "Y", "P", "A", "Z", se = "mom", cluster = "school")
```

## 8. How the estimates and standard errors are computed

This section is for readers who want to know exactly what runs.

**Point estimates are computed from formulas, not from `ivreg` or other
packages.** The IV estimate is the just-identified Wald ratio,
`cov(Z, Y) / cov(Z, A)` (weighted, and after partialling covariates when
present). The compass estimate residualizes `Z`, `Y`, and `P` on `A` to obtain
the loading ratio `delta_hat`, forms the corrected outcome `Y* = Y - delta_hat *
P`, and regresses `Y*` on `A`. Depending only on base R keeps the package light
and its behavior transparent.

**Standard errors come from a single method-of-moments (GMM) sandwich.** Three
moment conditions identify `delta_hat`, `tau_comp`, and `tau_IV` jointly; the
sandwich covariance of the three is computed once. The SE of `Delta` is the
square root of `a' V a` with `a = (0, -1, 1)`, and each **individual** SE is the
square root of the corresponding diagonal entry of the same `V`. Because they
share one covariance matrix, the individual SEs and the `Delta` SE are mutually
consistent. For `tau_IV` this SE equals the heteroskedasticity-robust (HC0)
standard error of a just-identified IV estimator.

When `cluster = ` is supplied, the same three moment conditions are summed
within each cluster before the sandwich is formed, so the "meat" of the
sandwich reflects between-cluster variation rather than between-observation
variation. This is a change to the covariance estimate only; the moment
conditions, and therefore the point estimates, are untouched.

For `Delta` you may instead request a **bootstrap** interval (`se =
"bootstrap"`), which resamples rows and recomputes `Delta`. The analytic and
bootstrap SEs agree up to Monte Carlo error. With `cluster = `, the bootstrap
resamples whole clusters instead of rows.

```{r bootstrap}
compare_iv_cv(ivc_example, "Y", "P", "A", "Z",
              se = "bootstrap", n_boot = 2000, seed = 1)
```

## 9. A caution: weak instruments

The diagnostic needs a **relevant** instrument. If `Z` barely moves `A`, the
correction becomes unstable and the standard errors blow up, so `Delta` will
look "not significant" simply because nothing can be estimated precisely. That
is not evidence of a valid instrument.

```{r weak}
d_weak <- ivc_simulate(n = 300, gamma = 0.05, kappa = 0.2, seed = 3)
compare_iv_cv(d_weak, "Y", "P", "A", "Z", se = "mom")
```

Notice the very wide interval. **Non-rejection is never proof of validity**; it
is only informative when the instrument is relevant. Check instrument strength
(e.g. the first-stage relationship between `Z` and `A`) before trusting a
non-rejection.

This is a different failure mode from a **weak compass loading**. Instrument
relevance is about `Z`'s relationship to `A`; the compass loading is about
`Z`'s relationship to `P` *conditional on* `A` -- the denominator of
`delta_hat`. If that conditional relationship is too weak to distinguish from
zero, `delta_hat` (and therefore `tau_comp`) becomes an unstable ratio even
when `Z` predicts `A` perfectly well. `estimate_cv()` checks this
automatically and warns:

```{r weak-loading}
d_noload <- data.frame(A = rbinom(500, 1, 0.5), Z = rnorm(500),
                       P = rnorm(500), Y = rnorm(500))  # Z unrelated to P given A
estimate_cv(d_noload, "Y", "P", "A", "Z")
```

Set `weak_loading_warn = 0` to silence this check, or raise the `|z|`
threshold (default `2`) to make it more conservative.

## 10. Choosing a compass variable: the conditional local independence test

Both `estimate_cv()` and `compare_iv_cv()` *assume* that `Z` is a valid
compass: that conditional on the treatment and the (unobserved) confounder,
`Z` is independent of both the pretest and the posttest. That assumption is
not directly testable with a single candidate variable. If you have **two or
more** candidate compass variables, `ivc_cli_test()` checks a testable
implication of that assumption -- a vanishing tetrad -- separately within each
treatment level, following Lee & Kim (2026, *Journal of Educational
Evaluation*, 39(1)).

```{r cli-sim}
set.seed(4)
n <- 2000
U  <- rnorm(n)
G  <- rbinom(n, 1, plogis(U))
P  <- 0.5 * U + rnorm(n)
Y  <- 0.5 * G + 0.5 * U + rnorm(n)
C1 <- 0.5 * U + rnorm(n)   # a plausible compass candidate
C2 <- 0.5 * U + rnorm(n)   # a second, independent candidate
d_cli <- data.frame(G, P, Y, C1, C2)

ivc_cli_test(d_cli, pretest = "P", posttest = "Y", treat = "G",
            compass = c("C1", "C2"))
```

A rejection means the four variables `(P, Y, C1, C2)` are jointly
inconsistent with the assumption -- but, as with any global specification
test, it cannot say *which* candidate is at fault. With three or more
candidates, every pair is tested; comparing the resulting `p_adj` values
across pairs is a reasonable way to screen for the pair with the weakest
evidence against the assumption before it is passed to `estimate_cv()`.

## 11. Planning a study

Before committing to an instrument, you can ask whether a given design could
even detect a violation, using `ivc_power()`.

```{r power}
ivc_power(n = 1000, gamma = 1, kappa = 0.2, n_rep = 100, se = "mom", seed = 7)
```

Set `kappa = 0` to read off the empirical Type I error rate instead. Note that
`ivc_simulate()` and `ivc_power()` are tools for **method checking and
teaching** -- they need the true violation strength `kappa`, which real data
never reveal. For an actual analysis you use `compare_iv_cv()` only.
