---
title: "Getting started with ivctools: diagnosing single-instrument exogeneity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ivctools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

## 1. When is this package for you?

`ivctools` is for a specific but common situation in education and the social
sciences: you have a **pretest--posttest design**, you want to estimate the
causal effect of a treatment, and you are relying on a **single instrumental
variable** whose exogeneity you cannot fully defend.

You should consider the IVC diagnostic if you can identify four things in your
data:

- an **outcome** measured after treatment (the posttest, `Y`),
- a **pretest** measured before treatment (`P`),
- a **treatment** indicator or dose (`A`),
- a candidate **instrument** (`Z`) that you hope affects `Y` only through `A`.

Standard overidentification tests (Sargan, Hansen J) cannot help here: they
need *two or more* instruments. The IVC diagnostic instead uses the pretest as
a built-in benchmark, so it works with a single instrument.

```{r setup}
library(ivctools)
```

## 2. The idea in one paragraph

If the instrument `Z` is valid, two different estimates of the treatment effect
should agree: the conventional IV estimate (`tau_IV`) and a *compass-corrected*
estimate built from the pretest (`tau_comp`). Their difference,
`Delta = tau_IV - tau_comp`, has expectation zero under a valid instrument. A
confidence interval for `Delta` that **excludes zero** is evidence that `Z`
violates exogeneity. Crucially, `tau_comp` remains a valid estimate of the
effect even when `Z` is invalid, so the gap between the two is informative.

## 3. A first analysis

The package ships with a small simulated dataset, `ivc_example`, which has a
mild exogeneity violation built in.

```{r data}
data(ivc_example)
head(ivc_example)
```

Run the diagnostic with one call. Give it the data frame and the column names.

```{r first-fit}
fit <- compare_iv_cv(ivc_example,
                     outcome = "Y", pretest = "P",
                     treat = "A", instrument = "Z",
                     se = "mom")
fit
```

## 4. Reading the output

Each quantity now reports its own standard error:

- **`tau_IV`** -- the conventional IV estimate. Biased if `Z` is invalid.
- **`tau_comp`** -- the compass-corrected estimate. Your best point estimate of
  the effect whether or not `Z` is valid.
- **`delta_hat`** -- the loading ratio used in the correction.
- **`Delta`** and its CI -- the diagnostic itself. If the CI excludes 0, the
  conclusion line flags evidence against exogeneity.

You can pull any value out directly; the fitted object is a plain list.

```{r extract}
fit$tau_comp
fit$se_tau_comp
fit$Delta
fit$ci
fit$reject
```

A picture often communicates the diagnostic best:

```{r plot}
plot(fit)
```

## 5. Individual estimates and their standard errors

If you only want one estimate with its standard error, use the single-estimate
functions with `se = TRUE`.

```{r individual}
estimate_iv(ivc_example, "Y", "P", "A", "Z", se = TRUE)
estimate_cv(ivc_example, "Y", "P", "A", "Z", return_delta = TRUE, se = TRUE)
```

## 6. Covariates and survey weights

Real data usually need controls and, for complex samples, weights. Supply them
by column name.

```{r covariates, eval = FALSE}
compare_iv_cv(my_data,
              outcome = "posttest", pretest = "pretest",
              treat = "catholic", instrument = "z",
              covariates = c("female", "ses", "region"),
              weights = "survey_weight",
              se = "mom")
```

Covariates are handled by Frisch--Waugh--Lovell partialling: the outcome,
pretest, treatment, and instrument are residualized on the covariates before
the estimators are applied. With neither `covariates` nor `weights`, the
results reduce **exactly** to the covariate-free forms.

## 7. How the estimates and standard errors are computed

This section is for readers who want to know exactly what runs.

**Point estimates are computed from formulas, not from `ivreg` or other
packages.** The IV estimate is the just-identified Wald ratio,
`cov(Z, Y) / cov(Z, A)` (weighted, and after partialling covariates when
present). The compass estimate residualizes `Z`, `Y`, and `P` on `A` to obtain
the loading ratio `delta_hat`, forms the corrected outcome `Y* = Y - delta_hat *
P`, and regresses `Y*` on `A`. Depending only on base R keeps the package light
and its behavior transparent.

**Standard errors come from a single method-of-moments (GMM) sandwich.** Three
moment conditions identify `delta_hat`, `tau_comp`, and `tau_IV` jointly; the
sandwich covariance of the three is computed once. The SE of `Delta` is the
square root of `a' V a` with `a = (0, -1, 1)`, and each **individual** SE is the
square root of the corresponding diagonal entry of the same `V`. Because they
share one covariance matrix, the individual SEs and the `Delta` SE are mutually
consistent. For `tau_IV` this SE equals the heteroskedasticity-robust (HC0)
standard error of a just-identified IV estimator.

For `Delta` you may instead request a **bootstrap** interval (`se =
"bootstrap"`), which resamples rows and recomputes `Delta`. The analytic and
bootstrap SEs agree up to Monte Carlo error.

```{r bootstrap}
compare_iv_cv(ivc_example, "Y", "P", "A", "Z",
              se = "bootstrap", n_boot = 2000, seed = 1)
```

## 8. A caution: weak instruments

The diagnostic needs a **relevant** instrument. If `Z` barely moves `A`, the
correction becomes unstable and the standard errors blow up, so `Delta` will
look "not significant" simply because nothing can be estimated precisely. That
is not evidence of a valid instrument.

```{r weak}
d_weak <- ivc_simulate(n = 300, gamma = 0.05, kappa = 0.2, seed = 3)
compare_iv_cv(d_weak, "Y", "P", "A", "Z", se = "mom")
```

Notice the very wide interval. **Non-rejection is never proof of validity**; it
is only informative when the instrument is relevant. Check instrument strength
(e.g. the first-stage relationship between `Z` and `A`) before trusting a
non-rejection.

## 9. Planning a study

Before committing to an instrument, you can ask whether a given design could
even detect a violation, using `ivc_power()`.

```{r power}
ivc_power(n = 1000, gamma = 1, kappa = 0.2, n_rep = 100, se = "mom", seed = 7)
```

Set `kappa = 0` to read off the empirical Type I error rate instead. Note that
`ivc_simulate()` and `ivc_power()` are tools for **method checking and
teaching** -- they need the true violation strength `kappa`, which real data
never reveal. For an actual analysis you use `compare_iv_cv()` only.
